Notes on filtration of logics containing {\bf K5}

### 2021-01-30 filtration-stuff/Text.tex
diagram sketch

\begin{theorem}
  ${\bf K}5$ does not admit strict filtrations.
\end{theorem}

@@

>>>>

\begin{theorem}\label{sad}
  ${\bf K}5$ does not admit strict filtrations.
\end{theorem}

@@

\section{``Finite'' canonical models}

Let $\mathcal{L}$ be a normal modal logic, $\mathcal{M}_{\mathcal{L}}$ its canonical model, and $\varphi$. Let us put $\Gamma = \operatorname{Sub}(\varphi) \cup \{ \neg \psi \: | \: \psi \in \operatorname{Sub}(\varphi) \}$.

A subset $\Delta \subseteq \Gamma$ is a \emph{finite $\mathcal{L}$-theory} if $\bigwedge \Delta \notin \mathcal{L}$. A subset $\Delta$ is maximal, if (the following are obviouslyequivalent):
\begin{enumerate}
  \item $\Delta$ is maximal amongst finite $\mathcal{L}$-theories,
  \item For each $\psi \in \operatorname{Sub}(\varphi)$ either $\psi \in \Delta$ or
  $\neg \psi \in \Delta$.
\end{enumerate}

Every finite $\mathcal{L}$-theory is clearly can be extended to some maximal one. It is the finite version of Lindenbaum's lemma.

\begin{defin}
  Let $\mathcal{L}$ be a normal modal logic and $\varphi \not\in \mathcal{L}$. A ``finite'' canonical model is a triple $\mathcal{M}_{\mathcal{L}}^{\varphi} = \langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi}, \vartheta_{\mathcal{L}}^{\varphi} \rangle$, where
  \begin{enumerate}
    \item $W_{\mathcal{L}}^{\varphi}$ is the set all maximal theories that extend finite $\mathcal{L}$-theories
    \item $\Delta_1 R \Delta_2$ iff $\Diamond \psi \in \Delta_2 \Rightarrow \psi \in \Delta_1$
    \item $\vartheta_{\mathcal{L}}^{\varphi}(p) = \{ \Delta \in W_{\mathcal{L}}^{\varphi} \: | \: p \in \Delta \}$.
  \end{enumerate}
\end{defin}

\begin{defin} Let $\varphi$ be a modal formula, $\mathcal{L}$ a logic that admits strict filtration, and $\mathcal{M}$ an $\mathcal{L}$-model that refutes $\varphi$. A \emph{filtration diagram} of a model $\widehat{\mathcal{M}}$ (a filtration of $\mathcal{M}$ through $\varphi$) is a formula $\operatorname{FDiagram}(\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)}) = \bigwedge \Gamma \wedge \bigwedge \Delta$, where
  \begin{itemize}
    \item $\Gamma =
    \{ \psi \in \operatorname{Sub}(\varphi) \: | \: \exists w \in \underline{\mathcal{M}} \: \widehat{M}, \hat{w} \models \psi \}$
    \item $\Delta = \{ \neg \psi \in \operatorname{Sub}(\varphi) \: | \: \psi \in \operatorname{Sub}(\varphi) \: \& \: \mathcal{M} \not\models \psi \}$
  \end{itemize}
\end{defin}

\begin{lemma}
  Let $\varphi$ be a modal formula, $\mathcal{L}$ a logic that admits strict filtration, and $\mathcal{M}$ an $\mathcal{L}$-model that refutes $\varphi$. Then $\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)} \models \operatorname{FDiagram}(\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)})$
\end{lemma}

\begin{defin}
  Let $\mathcal{M}$ be a model and $\Psi$ a set of formulas, then a $\Psi$-theory of a model, denoted as $\operatorname{Th}_{\Psi}$, is the set $\operatorname{Th}_{\Psi} = \{ \varphi \in \Psi \: | \: \mathcal{M} \models \varphi \}$.
\end{defin}

\begin{lemma}
  Let $\mathcal{L}$ be a modal logic that admits strict filtrations and $\varphi \not\in \mathcal{L}$.
  Let $\mathcal{M}$ be a model that refutes $\varphi$ and $\mathcal{N}$ be a different model such that $\mathcal{N} \models \mathcal{L}$. If $\mathcal{N} \models \operatorname{FDiagram}(\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)})$, then $\operatorname{Th}_{\operatorname{Sub}(\varphi)}(\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)}) = \operatorname{Th}_{\operatorname{Sub}(\varphi)}(\mathcal{N})$. In particular, $\mathcal{M}$ refutes $\varphi$.
\end{lemma}

\begin{theorem} Let $\mathcal{L}$ be a complete normal modal logic, then the following are equivalent:
  \begin{enumerate}
    \item $\mathcal{L}$ admits strict filtrations
    \item If $\varphi \not\in \mathcal{L}$, then the ``finite'' canonical model $\mathcal{M}_{\mathcal{L}}^{\varphi}$ refutes $\varphi$
  \end{enumerate}
\end{theorem}

  \begin{enumerate}
    \item $(\Rightarrow)$

    Let $\varphi \not\in \mathcal{L}$, there exists a model $\mathcal{M}$ that refutes $\varphi$, so there exists $w \in \underline{\mathcal{M}}$ such that $\mathcal{M}, w \not\models \varphi$. But $\mathcal{L}$ admits strict filtration, so there exists a model $\widehat{\mathcal{M}} = \langle \widehat{W}, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{W} = W / \sim_{\varphi}$ and $\hat{R}^{min} \subseteq \widehat{R} \subseteq \hat{R}^{max}$. In particular, $\widehat{\mathcal{M}}, \widehat{w} \not\models \varphi$.

    By lemma above, $\widehat{\mathcal{M}} \models \operatorname{FDiagram}(\widehat{\mathcal{M}})$

    Consider a ``finite'' canonical model
    $\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)} = \langle W_{\mathcal{L}}^{\operatorname{Sub}(\varphi)}, R_{\mathcal{L}}^{\operatorname{Sub}(\varphi)}, \vartheta_{\mathcal{L}}^{\operatorname{Sub}(\varphi)} \rangle$.
    Let us show that $\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}}$ refutes $\varphi$. For that, we show that $\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)} \models \operatorname{FDiagram}(\widehat{\mathcal{M}})$.

    \item $(\Leftarrow)$

    Let $\mathcal{M}_{\mathcal{L}}^{\varphi} = $ be a ``finite'' canonical model that refutes $\varphi$.

    Consider $W_{\mathcal{L}}^{\varphi} / \sim_{\operatorname{Sub}(\varphi)}$. Clearly $W_{\mathcal{L}}^{\varphi} = W_{\mathcal{L}}^{\varphi} / \sim_{\operatorname{Sub}(\varphi)}$ since

    \begin{center}
      $\Delta_1 \sim_{\operatorname{Sub}(\varphi)} \Delta_2$ iff $\forall \psi \in \operatorname{Sub}(\varphi) \:\: W_{\mathcal{L}}^{\varphi}, \Delta_1 \models \psi \Leftrightarrow W_{\mathcal{L}}^{\varphi}, \Delta_2 \models \psi$ iff $\Delta_1 = \Delta_2$.
    \end{center}

    $R_{\mathcal{L}}^{\varphi}$ also trivially satisfies the required defined. Therefore, $\mathcal{L}$ admits strict filtrations since $\mathcal{M}_{\mathcal{L}}^{\varphi}$ is already finite. In particular, ${\bf K}5$ does not have the ``finite'' canonical property that follows from the contraposition of this statement and Theorem~\ref{sad}.

  \end{enumerate}
\end{proof}
